¥documentclass[11pt]{article}

¥usepackage[margin=1in]{geometry}
¥usepackage{amsmath,amssymb,amsthm}
¥usepackage{enumitem}
¥usepackage{hyperref}
¥usepackage{xcolor}
¥usepackage{booktabs}

¥hypersetup{colorlinks=true, linkcolor=blue, urlcolor=blue, citecolor=blue}

¥title{TOAE209/TOAH209 -- Design and Analysis of Algorithms¥¥[4pt]
¥large Week 6: Theoretical Questions}
¥author{Foreign Trade University}
¥date{}

¥begin{document}
¥maketitle

¥section*{Instructions}

Part A contains 17 questions requiring written explanations, short proofs, or
recurrence-solving. Part B is a 10-question quiz (true/false and multiple choice). Attempt
every question on your own before consulting Part C (Solutions) at the end of this document.

¥section*{Part A: Theory Questions}

¥begin{enumerate}[label=¥textbf{A¥arabic*.}, leftmargin=2.2em]

¥item Describe the ¥emph{divide-and-conquer} paradigm in general terms. Name and explain
its three steps, and write the general form of the running-time recurrence that results.

¥item State the ¥textsc{Merge} subroutine for two sorted lists and prove that it produces a
sorted list containing exactly the elements of both inputs. State a loop invariant and use
it in your proof.

¥item Prove that ¥textsc{Mergesort} correctly sorts any input array, by strong induction on
the array length.

¥item Write down the recurrence for mergesort's running time and solve it by
¥emph{unrolling} (repeated substitution). Show the intermediate steps and identify the
number of levels.

¥item Solve the same mergesort recurrence $T(n) = 2T(n/2) + cn$ using a ¥emph{recursion
tree}. For each level $k$, give the number of nodes, the size of each node, and the total
work at that level; then sum over all levels.

¥item State the ¥textbf{Master Theorem} in full, including all three cases and the
condition (comparison of $d$ with $¥log_b a$) that selects each case.

¥item For each of the following recurrences, apply the Master Theorem and give the
$¥Theta$-class: (a) $T(n) = 2T(n/2) + ¥Theta(n)$; (b) $T(n) = T(n/2) + ¥Theta(1)$; (c)
$T(n) = 3T(n/2) + ¥Theta(n)$; (d) $T(n) = 4T(n/2) + ¥Theta(n)$; (e) $T(n) = 2T(n/2) +
¥Theta(n^2)$. State which case applies and why.

¥item For each recurrence, give the $¥Theta$-class via the Master Theorem and identify the
case: (a) $T(n) = 9T(n/3) + ¥Theta(n)$; (b) $T(n) = T(n/2) + ¥Theta(n)$; (c) $T(n) =
8T(n/2) + ¥Theta(n^3)$; (d) $T(n) = 7T(n/2) + ¥Theta(n^2)$.

¥item Define an ¥emph{inversion} of an array. How many inversions does a sorted array of
length $n$ have? How many does a strictly reverse-sorted array of length $n$ have? Justify
the reverse-sorted count.

¥item Explain how counting inversions can be done in $O(n ¥log n)$ by modifying mergesort.
In particular, when merging sorted halves $L$ and $R$ and the algorithm emits an element of
$R$ before exhausting $L$, exactly how many inversions does that single step account for,
and why?

¥item Write down and justify the recurrence for the inversion-counting algorithm. Why is it
the same order as mergesort itself?

¥item State the recurrence for binary search and solve it. Then explain why binary search
is "divide and conquer with $a = 1$" and what that means for the shape of its recursion
tree (compared to mergesort's).

¥item Describe the divide-and-conquer algorithm for the maximum-subarray problem. In
particular, describe the three cases the maximum subarray can fall into, and how the
¥emph{cross-boundary} case is computed in $O(n)$. State and solve the recurrence.

¥item Explain exponentiation by squaring. Write the recursive identity for $x^n$ (even and
odd cases), state the recurrence for the number of multiplications, solve it, and explain
why computing $x^{n/2}$ ¥emph{once} (rather than twice) is essential to achieving
$¥Theta(¥log n)$.

¥item Describe ¥emph{quickselect} for finding the $k$-th smallest element. Explain why, with
a random pivot, the ¥emph{expected} running time is $¥Theta(n)$, and why the Master Theorem
does not directly apply to its recurrence. What is the worst-case running time, and when
does it occur?

¥item Explain the divide-and-conquer algorithm for finding a ¥emph{majority element} (a
value occurring more than $n/2$ times). Why does it suffice to consider only the majority
candidates of the two halves in the combine step? State and solve the recurrence.

¥item Karatsuba multiplication satisfies $T(n) = 3T(n/2) + ¥Theta(n)$, while the naive
divide-and-conquer multiplication satisfies $T(n) = 4T(n/2) + ¥Theta(n)$. Solve both
recurrences with the Master Theorem and explain precisely why reducing the number of
recursive multiplications from $4$ to $3$ changes the asymptotic running time (and why
shaving the $¥Theta(n)$ combine term would ¥emph{not} help).

¥end{enumerate}

¥newpage
¥section*{Part B: Quiz}

¥begin{enumerate}[label=¥textbf{B¥arabic*.}, leftmargin=2.2em]

¥item ¥textbf{(Multiple choice)} The recurrence for mergesort is:
¥begin{enumerate}[label=(¥alph*)]
    ¥item $T(n) = T(n-1) + ¥Theta(n)$
    ¥item $T(n) = 2T(n/2) + ¥Theta(n)$
    ¥item $T(n) = 2T(n/2) + ¥Theta(1)$
    ¥item $T(n) = T(n/2) + ¥Theta(n)$
¥end{enumerate}

¥item ¥textbf{(True/False)} Merging two sorted lists of total length $n$ can be done in
$O(n)$ time.

¥item ¥textbf{(Multiple choice)} By the Master Theorem, $T(n) = 2T(n/2) + ¥Theta(n)$ solves
to:
¥begin{enumerate}[label=(¥alph*)]
    ¥item $¥Theta(n)$
    ¥item $¥Theta(n ¥log n)$
    ¥item $¥Theta(n^2)$
    ¥item $¥Theta(¥log n)$
¥end{enumerate}

¥item ¥textbf{(Multiple choice)} For $T(n) = a T(n/b) + ¥Theta(n^d)$, Case 1 of the Master
Theorem (where $T(n) = ¥Theta(n^{¥log_b a})$) applies when:
¥begin{enumerate}[label=(¥alph*)]
    ¥item $d > ¥log_b a$
    ¥item $d = ¥log_b a$
    ¥item $d < ¥log_b a$
    ¥item $a = b$
¥end{enumerate}

¥item ¥textbf{(Short answer)} How many inversions does the array $[3, 2, 1]$ have? List
them.

¥item ¥textbf{(True/False)} Counting the inversions of an $n$-element array requires
$¥Theta(n^2)$ time in the worst case; no faster algorithm is possible.

¥item ¥textbf{(Multiple choice)} The number of multiplications used by exponentiation by
squaring to compute $x^n$ is:
¥begin{enumerate}[label=(¥alph*)]
    ¥item $¥Theta(n)$
    ¥item $¥Theta(¥log n)$
    ¥item $¥Theta(n ¥log n)$
    ¥item $¥Theta(¥sqrt{n})$
¥end{enumerate}

¥item ¥textbf{(Multiple choice)} The ¥emph{expected} running time of quickselect with a
random pivot is:
¥begin{enumerate}[label=(¥alph*)]
    ¥item $¥Theta(¥log n)$
    ¥item $¥Theta(n)$
    ¥item $¥Theta(n ¥log n)$
    ¥item $¥Theta(n^2)$
¥end{enumerate}

¥item ¥textbf{(True/False)} The divide-and-conquer maximum-subarray algorithm runs in
$¥Theta(n ¥log n)$, while Kadane's algorithm solves the same problem in $¥Theta(n)$.

¥item ¥textbf{(Short answer)} Karatsuba's recurrence is $T(n) = 3T(n/2) + ¥Theta(n)$. Give
its $¥Theta$-class (a numeric exponent to three decimal places is fine).

¥end{enumerate}

¥newpage
¥section*{Part C: Solutions}

¥subsection*{Part A}

¥begin{enumerate}[label=¥textbf{A¥arabic*.}, leftmargin=2.2em]

¥item ¥textbf{Divide and conquer} solves a problem by reducing it to smaller instances of
the same problem. Its three steps are: ¥textbf{Divide} -- split the size-$n$ input into $a$
subproblems each of size about $n/b$; ¥textbf{Conquer} -- solve each subproblem recursively
(stopping at small base cases); ¥textbf{Combine} -- merge the subproblem solutions into a
solution for the original instance, doing some additional work. The running time satisfies a
recurrence
¥[
    T(n) = a¥,T(n/b) + (¥text{divide + combine cost}),
¥]
typically written $T(n) = a¥,T(n/b) + ¥Theta(n^d)$.

¥item ¥textsc{Merge}$(L,R)$ keeps pointers $i,j$ into the two sorted lists and repeatedly
appends $¥min(L[i], R[j])$ to the output $M$, advancing the corresponding pointer; when one
list is exhausted, it appends the remaining tail of the other. ¥textit{Loop invariant:} at
the start of each iteration, (i) $M$ is sorted, and (ii) every element already in $M$ is
$¥le$ every element remaining in $L[i¥mathbin{:}]$ and $R[j¥mathbin{:}]$. ¥textit{Proof:}
initially $M$ is empty, so the invariant holds vacuously. Each iteration appends the smaller
of $L[i], R[j]$, which (because $L,R$ are sorted) is the smallest of all remaining elements,
hence $¥le$ everything still remaining and (by the invariant) $¥ge$ everything already in
$M$; so $M$ stays sorted and the invariant is preserved. When the loop ends, the remaining
tail of one list is internally sorted and all its elements are $¥ge$ everything in $M$, so
appending it leaves $M$ sorted. Each input element is appended exactly once, so $M$ contains
exactly the multiset union.

¥item ¥textit{Strong induction on $n = |A|$.} ¥textbf{Base case:} $n ¥le 1$; a list of
length $0$ or $1$ is already sorted and returned unchanged. ¥textbf{Inductive step:} for $n
¥ge 2$, both halves have length $< n$, so by the inductive hypothesis the recursive calls
return them sorted. By A2, ¥textsc{Merge} of two sorted halves returns a sorted permutation
of their union, which is $A$. Hence ¥textsc{Mergesort}$(A)$ is correct.

¥item Unrolling $T(n) = 2T(n/2) + cn$:
¥[
¥begin{aligned}
T(n) &= 2T(n/2) + cn = 2¥bigl(2T(n/4) + c¥tfrac n2¥bigr) + cn = 4T(n/4) + 2cn ¥¥
     &= 8T(n/8) + 3cn = ¥cdots = 2^k T(n/2^k) + k¥,cn.
¥end{aligned}
¥]
The recursion stops when $n/2^k = 1$, i.e.¥ $k = ¥log_2 n$, giving $2^k T(1) = n¥cdot O(1) =
O(n)$ plus accumulated combine cost $k¥,cn = cn¥log_2 n$. Therefore $T(n) = ¥Theta(n ¥log
n)$. There are $¥log_2 n + 1$ levels.

¥item Recursion tree for $T(n) = 2T(n/2) + cn$: level $k$ (root is level $0$) has $2^k$
nodes, each of size $n/2^k$, each doing $c(n/2^k)$ combine work. Total work at level $k$ is
$2^k ¥cdot c(n/2^k) = cn$ -- the ¥emph{same} at every level. There are $¥log_2 n + 1$
levels, so the total is $cn(¥log_2 n + 1) = ¥Theta(n ¥log n)$.

¥item ¥textbf{Master Theorem.} For $T(n) = a¥,T(n/b) + ¥Theta(n^d)$ with $a ¥ge 1$, $b > 1$,
$d ¥ge 0$:
¥begin{itemize}
    ¥item ¥textbf{Case 1} ($d < ¥log_b a$, i.e.¥ $a > b^d$): $T(n) = ¥Theta(n^{¥log_b a})$ --
    leaves dominate.
    ¥item ¥textbf{Case 2} ($d = ¥log_b a$, i.e.¥ $a = b^d$): $T(n) = ¥Theta(n^d ¥log n)$ --
    every level does equal work.
    ¥item ¥textbf{Case 3} ($d > ¥log_b a$, i.e.¥ $a < b^d$): $T(n) = ¥Theta(n^d)$ -- the root
    dominates.
¥end{itemize}

¥item (a) $a=2,b=2,d=1$, $¥log_2 2 = 1 = d$ $¥Rightarrow$ Case 2, $¥Theta(n ¥log n)$. (b)
$a=1,b=2,d=0$, $¥log_2 1 = 0 = d$ $¥Rightarrow$ Case 2, $¥Theta(¥log n)$. (c) $a=3,b=2,d=1$,
$¥log_2 3 ¥approx 1.585 > 1$ $¥Rightarrow$ Case 1, $¥Theta(n^{¥log_2 3}) ¥approx
¥Theta(n^{1.585})$. (d) $a=4,b=2,d=1$, $¥log_2 4 = 2 > 1$ $¥Rightarrow$ Case 1,
$¥Theta(n^2)$. (e) $a=2,b=2,d=2$, $¥log_2 2 = 1 < 2$ $¥Rightarrow$ Case 3, $¥Theta(n^2)$.

¥item (a) $a=9,b=3,d=1$, $¥log_3 9 = 2 > 1$ $¥Rightarrow$ Case 1, $¥Theta(n^2)$. (b)
$a=1,b=2,d=1$, $¥log_2 1 = 0 < 1$ $¥Rightarrow$ Case 3, $¥Theta(n)$. (c) $a=8,b=2,d=3$,
$¥log_2 8 = 3 = d$ $¥Rightarrow$ Case 2, $¥Theta(n^3 ¥log n)$. (d) $a=7,b=2,d=2$, $¥log_2 7
¥approx 2.807 > 2$ $¥Rightarrow$ Case 1, $¥Theta(n^{¥log_2 7}) ¥approx ¥Theta(n^{2.807})$.

¥item An ¥textbf{inversion} of an array $A$ is a pair of indices $(i,j)$ with $i < j$ but
$A[i] > A[j]$. A sorted (non-decreasing) array has ¥textbf{0} inversions (no pair is out of
order). A strictly reverse-sorted array of length $n$ has ¥textbf{$¥binom{n}{2} =
¥frac{n(n-1)}{2}$} inversions: ¥emph{every} one of the $¥binom{n}{2}$ pairs $i < j$ has
$A[i] > A[j]$, which is the maximum possible.

¥item Run mergesort; whenever ¥textsc{Merge} emits an element $R[j]$ from the right half
while $L$ still has $|L| - i$ elements remaining (i.e.¥ $R[j] < L[i]$, so the algorithm
takes from the right), those $|L| - i$ remaining left elements are each ¥emph{to the left}
of $R[j]$ in the original array yet ¥emph{larger} than $R[j]$ -- so each forms exactly one
inversion with $R[j]$. Thus a single such step accounts for exactly $|L| - i$ inversions.
Summed across all merges, this counts every cross-inversion exactly once; combined with the
inversions counted recursively inside each half, the total is the array's inversion count.

¥item $T(n) = 2T(n/2) + ¥Theta(n)$, identical to mergesort's, because the only addition to
¥textsc{Merge} is incrementing a counter by $|L| - i$ at each right-emission step -- $O(1)$
extra work per merge step, leaving the per-merge cost $¥Theta(n)$ unchanged. By the Master
Theorem (Case 2), $T(n) = ¥Theta(n ¥log n)$.

¥item Binary search: $T(n) = T(n/2) + ¥Theta(1)$. By the Master Theorem with $a=1$, $b=2$,
$d=0$ ($¥log_2 1 = 0 = d$, Case 2), $T(n) = ¥Theta(¥log n)$. It is "divide and conquer with
$a=1$" because, although the input is split into two halves, the algorithm ¥emph{recurses
into only one} of them -- so the recursion tree is a single ¥emph{path} of length $¥log_2 n$
(not a branching tree as in mergesort), and the total work is just $¥Theta(1)$ per level
times $¥log_2 n$ levels.

¥item Split $A$ at the midpoint $m$. The maximum-sum subarray is one of: (i) entirely in the
left half, (ii) entirely in the right half, or (iii) ¥emph{crossing} the midpoint. Cases (i)
and (ii) are solved by recursion. For (iii), the best crossing subarray is the best suffix
of the left half (computed by scanning leftward from $m$ keeping a running maximum) joined to
the best prefix of the right half (scanning rightward) -- $O(n)$ total. Return the max of the
three. The recurrence is $T(n) = 2T(n/2) + ¥Theta(n) = ¥Theta(n ¥log n)$ (Master Theorem,
Case 2).

¥item Exponentiation by squaring uses
¥[
    x^n = ¥begin{cases} 1 & n=0,¥¥ (x^{n/2})^2 & n ¥text{ even},¥¥ x¥cdot(x^{(n-1)/2})^2 & n ¥text{ odd}.¥end{cases}
¥]
Letting $M(n)$ be the number of multiplications, $M(n) = M(¥lfloor n/2 ¥rfloor) + O(1)$
(one or two multiplications to square and possibly multiply by $x$). By the Master Theorem
($a=1,b=2,d=0$, Case 2), $M(n) = ¥Theta(¥log n)$. The key is that $x^{n/2}$ is computed
¥emph{once} and squared; if instead we wrote $x^n = x^{n/2}¥cdot x^{n/2}$ and computed each
factor separately, the recurrence would become $M(n) = 2M(n/2) + O(1) = ¥Theta(n)$ -- no
savings over naive repeated multiplication.

¥item ¥textbf{Quickselect}: choose a pivot $p$, partition $A$ into elements $<p$, $=p$, and
$>p$; if the $k$-th smallest lies in the $<p$ part, recurse there; if it is among the $=p$
part, return $p$; otherwise recurse on the $>p$ part with $k$ adjusted by the number of
discarded smaller elements. With a ¥emph{random} pivot, with constant probability the pivot
discards a constant fraction of the array, so the expected work satisfies $T(n) ¥le T(cn) +
O(n)$ for some constant $c < 1$ (e.g.¥ $c = 3/4$), a decreasing geometric series summing to
$O(n)$. The Master Theorem does not apply because (a) only ¥emph{one} subproblem is solved
and (b) its size is a constant ¥emph{fraction} of $n$ rather than a fixed $n/b$ for both
branches. The worst case is $¥Theta(n^2)$, occurring when the pivot is repeatedly the
largest or smallest remaining element (e.g.¥ a sorted input with a last-element pivot), so
each call shrinks the problem by only one.

¥item ¥textbf{Majority element via D¥&C}: split into halves; recursively find each half's
majority candidate. A value that is a majority of the ¥emph{whole} (occurs $> n/2$ times)
must be a majority of at least one half -- because if it were a strict minority in
¥emph{both} halves (occurring $¥le$ half the time in each), it would occur $¥le n/2$ times
overall, a contradiction. So the only candidates are the two halves' majority candidates;
the combine step counts each candidate's occurrences across the whole range (an $O(n)$
scan) and returns whichever exceeds $n/2$. Recurrence: $T(n) = 2T(n/2) + ¥Theta(n) =
¥Theta(n ¥log n)$.

¥item Karatsuba: $a=3,b=2,d=1$; $¥log_2 3 ¥approx 1.585 > 1$ $¥Rightarrow$ Case 1, $T(n) =
¥Theta(n^{¥log_2 3}) ¥approx ¥Theta(n^{1.585})$. Naive D¥&C: $a=4,b=2,d=1$; $¥log_2 4 = 2 >
1$ $¥Rightarrow$ Case 1, $T(n) = ¥Theta(n^2)$. Both are in Case 1, so the running time is
$¥Theta(n^{¥log_b a})$, governed ¥emph{entirely} by $a$ (the number of recursive
multiplications): dropping $a$ from $4$ to $3$ drops the exponent from $2$ to $¥log_2 3
¥approx 1.585$. Shaving the $¥Theta(n)$ combine term would not help, because in Case 1 the
combine term is asymptotically dominated by the leaf work $n^{¥log_b a}$ -- the running time
does not depend on $d$ at all once $d < ¥log_b a$.

¥end{enumerate}

¥subsection*{Part B}

¥begin{enumerate}[label=¥textbf{B¥arabic*.}, leftmargin=2.2em]

¥item ¥textbf{(b)} $T(n) = 2T(n/2) + ¥Theta(n)$: two recursive sorts of half-size halves,
plus a linear merge.

¥item ¥textbf{True.} Each comparison emits one element, so the total number of
comparisons/appends is at most the total length $n$; merging is $¥Theta(n)$.

¥item ¥textbf{(b)} $¥Theta(n ¥log n)$. ($a=b=2$, $d=1$, $¥log_2 2 = 1 = d$, Master Theorem
Case 2.)

¥item ¥textbf{(c)} $d < ¥log_b a$ (equivalently $a > b^d$): leaves dominate, $T(n) =
¥Theta(n^{¥log_b a})$.

¥item ¥textbf{3 inversions:} $(0,1)$ since $3>2$, $(0,2)$ since $3>1$, and $(1,2)$ since
$2>1$. This is the maximum $¥binom{3}{2} = 3$ for a length-3 array.

¥item ¥textbf{False.} Counting inversions can be done in $¥Theta(n ¥log n)$ via the modified
mergesort; the $¥Theta(n^2)$ brute force is not optimal.

¥item ¥textbf{(b)} $¥Theta(¥log n)$ -- each step halves the exponent.

¥item ¥textbf{(b)} $¥Theta(n)$ expected (with a random pivot). The worst case is
$¥Theta(n^2)$.

¥item ¥textbf{True.} The D¥&C version is $¥Theta(n ¥log n)$ (recurrence $2T(n/2) +
¥Theta(n)$); Kadane's single-pass dynamic program is $¥Theta(n)$.

¥item $¥Theta(n^{¥log_2 3}) ¥approx ¥Theta(n^{1.585})$.

¥end{enumerate}

¥end{document}
